---
title: Exploring Female Names in the U.S.
output:
  html_document:
    toc: true
---
  
A quick review of U.S. Female names in the database using dplyr, ggvis and dygraphs. 

I also created a shiny app. It is fun, take a look: http://shiny.journeysteps.com/namesDBexp/

Revisions: 

* Feb 27, 2016 : Adding dygraphs for interactive visualization

  
```{r message=FALSE, warning=FALSE}
library(dplyr) # to manipulate data
library(ggvis) # for data visualization
library(stringdist) # to look for similar names
library(data.table) # to load data faster
library(dygraphs) # for interactive data visualization
```
Loading data...
```{r warning=FALSE}
ndb <- tbl_df(read.csv("../input/NationalNames.csv")) #Kaggle
```
## Number of unique names by year and gender
```{r warning=FALSE}
ABY <- ndb %>%
  group_by(Year, Gender) %>%
  summarise(name_cnt = sum(Count),  # total count by Year and Gender
            unq_names = n_distinct(Name)) #count # distinct names by Year and Gender

ABY %>% 
  ggvis(~Year, ~unq_names, stroke=~factor(Gender)) %>% 
  layer_lines()
```
Comments

* Looks like people are more creative with female names than male names
* There is some explosion around 1915, probably due to immigration
* Since 1944 (end of WW2) till around 2010 (recent recession) the influx of original names has been strong

***
## Adding some basic stats and run a few queries
  
* name_cnt : number of times the name was used 
* nyears : number of years the name was used
* fy : first year of use (in DB)
* ly : last year of use (in DB)
* std_cnt : standard deviation of the yearly counts
* avg_cnt : average of the yearly counts
* nsd : normalized standard deviation

```{r warning=FALSE}

ABN <- ndb %>%
  filter(Gender == 'F') %>%
  group_by(Name) %>%
  summarise(name_cnt = sum(Count), 
            nyears = n_distinct(Year), 
            fy = min(Year), 
            ly = max(Year),
            std_cnt = sd(Count),
            avg_cnt = mean(Count),
            nsd = std_cnt/avg_cnt) 

# get peak year by name             
PeakYearF <- ndb %>%
  filter(Gender == 'F') %>%
  group_by(Name) %>%
  filter(min_rank(desc(Count * 10000 + Year)) ==1)  %>%  #workaround to sort by Count and Year 
  select ( Name, topYear = Year, CntTopYear = Count)
#merge data back into ABN 
ABN <- ABN %>% inner_join(PeakYearF, by = "Name") 

sample_n(ABN,5)
```
Using ABN and ndb, we can do several interesting queries. 

***
  
### Top 5 female names in database by use 
  
```{r warning=FALSE}
Top5F <- ABN  %>%
  ungroup() %>% # dplyr keeps prior grouping, here we need to remove it 
  arrange(desc(name_cnt)) %>%
  slice(1:5) #get top 5 

ndb   %>%
  filter(Gender == 'F', Name %in% Top5F$Name )  %>%
  select (Name, Year, Count) %>% 
  ggvis(~Year, ~Count, stroke = ~factor(Name)) %>% 
  layer_lines()
```
Even though these are the most popular names, their peak is in the past for most of them.
***
  
### Top female names that have been in use for 100 years and exceeded 1M uses
  
```{r warning=FALSE}
tmp3 <- ABN %>%
  filter(name_cnt > 1000000, nyears >= 100) %>%
  arrange(desc(name_cnt))

ndb   %>%
  filter(Gender == 'F', Name %in% tmp3$Name)  %>%
  select (Name, Year, Count) %>% 
  ggvis(~Year, ~Count, stroke = ~factor(Name)) %>% 
  layer_lines()

```

### Names that have recently peaked 

```{r warning=FALSE}
tmp2 <- ABN %>%
  filter(CntTopYear > 8000, topYear >= 2010, name_cnt > 50000) %>%
  arrange(desc(name_cnt))

ndb   %>%
  filter(Gender == 'F', Name %in% tmp2$Name)  %>%
  select (Name, Year, Count) %>% 
  ggvis(~Year, ~Count, stroke = ~factor(Name)) %>% 
  layer_lines()
```

### New names since 2008 that passed the 100 count nationally 

```{r warning=FALSE}
tmp4 <- ABN %>%
  filter( fy >= 2008,  CntTopYear> 100) %>%
  arrange(desc(name_cnt)) %>%
  slice(1:6)

ndb   %>%
  filter(Gender == 'F', Name %in% tmp4$Name)  %>%
  select (Name, Year, Count) %>% 
  ggvis(~Year, ~Count, stroke = ~factor(Name)) %>% 
  layer_lines()
```

### Old names that haven't been used since 2000 

```{r warning=FALSE}
tmp5 <- ABN %>%
  filter( fy < 1930, ly < 2000, CntTopYear> 500) %>%
  arrange(desc(name_cnt)) %>%
  slice(1:6)

ndb   %>%
  filter(Gender == 'F', Name %in% tmp5$Name)  %>%
  select (Name, Year, Count) %>% 
  ggvis(~Year, ~Count, stroke = ~factor(Name)) %>% 
  layer_lines()

```

### "One Shot" names, had a one time high peak 

The most recent three might not qualify since their peak is quite recent.

```{r warning=FALSE}
tmp7 <- ABN %>%
  filter(  name_cnt > 1000, 
           nsd >= 2 # using normalized std dev. to get names with high peaks 
           ) %>%
  arrange(desc(name_cnt)) %>%
  slice(1:6)

ndb   %>%
  filter(Gender == 'F', Name %in% tmp7$Name )  %>%
  select (Name, Year, Count) %>% 
  ggvis(~Year, ~Count, stroke = ~factor(Name)) %>% 
  layer_lines()

```
***

## Consistent names, ploting with dygraphs (NEW)

Filtering names with low variability and top year less than 2000 to exclude a couple of names that have higher volume and make the rest too flat. 

```{r warning=FALSE, message=FALSE, fig.width=9}
tmp8 <- ABN %>%
  filter(CntTopYear < 2000,
         nsd < 0.5) %>%   # using normalized std dev. to get names with less count variability in time
  arrange(desc(name_cnt)) %>%
  slice(1:10) # increase to 10 since highlighting makes easier to read the graph

library(lubridate)
library(tidyr)
library(xts)

nameSeries <-ndb   %>%
  filter(Gender == 'F', Name %in% tmp8$Name )  %>%
  select (Name, Year  , Count) %>%
  mutate( Year =ymd(paste0(Year,'0101') ))  #convert Year Integer to date

# re-shape table to pivot names into columns
nameSeries <- spread(nameSeries, Name, Count) 

#convert to time series data class
vis <-xts(nameSeries, order.by = nameSeries$Year)

#plot with dygraph to make it interactive
# add range selector, allows to change the range of years for the visualization
# add highlighting of a particular name
dygraph(vis) %>% 
  dyRangeSelector() %>%
  dyHighlight(highlightCircleSize = 4, 
              highlightSeriesBackgroundAlpha = 0.5,
              hideOnMouseOut = TRUE)%>%
  dyLegend(show = "follow")

```

***
  
## Looking for similar names using string distance
  
Use stringdist to find similar names, not perfect but decent close enough names.

```{r warning=FALSE}
# reducing the list of names to more recent and familar names 
# names used at least till 1961 and that had a top count higher than 200 
relativePopFNames <- ABN %>% 
  filter(ly>1960,CntTopYear>200)  %>%
  select(Name) 
vnames <-relativePopFNames$Name

```
### Carla

```{r warning=FALSE}
vdist<- stringdist("carla",tolower(vnames),method="osa") 
filteredList <- vnames[vdist <=1 ]

snames <- ABN %>%
  filter( Name %in% filteredList) %>%
  arrange(desc(name_cnt)) %>%
  slice(1:6)

ndb   %>%
  filter(Gender == 'F', Name  %in% snames$Name )  %>%
  select (Name, Year, Count) %>% 
  ggvis(~Year, ~Count, stroke = ~factor(Name)) %>% 
  layer_lines()

```

### Tania

```{r warning=FALSE}

vdist<- stringdist("tania",tolower(vnames),method="osa") 
filteredList <- vnames[vdist <=1 ]

snames <- ABN %>%
  filter( Name %in% filteredList) %>%
  arrange(desc(name_cnt)) %>%
  slice(1:6)

ndb   %>%
  filter(Gender == 'F', Name  %in% snames$Name )  %>%
  select (Name, Year, Count) %>% 
  ggvis(~Year, ~Count, stroke = ~factor(Name)) %>% 
  layer_lines()


```